Forward Sensitivity Equations in the Presence of Events

### 2019-01-25 paper.tex
Some minor edits

% Write abstract here
\begin{abstract}
	Forward sensitivity equations are frequently used in optimization. Based on forward sensitivities, the gradient and Hessian of the least squares function can be derived which allow to use gradient-based optimization methods. In the presence of events, i.e. sudden changes of the state variables, the differential equations and corresponding sensitivity equations are structurally unchanged. However, the events on states need to be accounted for as events in the sensitivities. Here, we derive these necessary events and present them in a way that helps with the implementation in computational software. Finally, the sensitivity events are illustrated on a typical example from the field of pharmacometrics.
\end{abstract}

@@

% Main Part
\section{Main Part}
\label{sec:main}
Let $\dot x = f(x, p, t)$ be a system of ordinary differential equations (ODEs) with the states $x(t) \in \mathbb R^n$, parameters $p \in \mathbb R^m$ and time $t \in \mathbb R$. The sensitivity equations corresponding to this dynamic system are
\begin{align}
\begin{aligned}
    \frac{d}{dt}\frac{\partial x_i}{\partial p_j} &= \frac{d}{dp_j} f_i(x, p, t)\\
        & = \frac{\partial f_i}{\partial x_k}\frac{\partial x_k}{\partial p_j} +
            \frac{\partial f_i}{\partial p_j}.
\end{aligned}
 \label{eq:sens}
\end{align}
Let us assume that an event occurs at time $t_e$ that changes the current state vector $x(t_e)$ to the value $v_e \in \mathbb R^n$. Both $t_e$ and $v_e$ are assumed to be parameters for which sensitivities are to be determined.

@@

The derivation is based on linearization of the ODE around $x(t_e) = x_e$, this is the value of $x(t)$ at the latest time point before the event occurs. With this choice, the linearized ODE reads
\begin{align}
    \dot x \doteq A(p, t) x + b(p, t)
    \label{eq:linearized}
\end{align}
with
\begin{align*}
A(p, t) &= \left.\frac{\partial f}{\partial x}\right|_{x_e}(p, t)\\
b(p, t) &= f(x_e, p, t)- \left.\frac{\partial f}{\partial x}\right|_{x_e}(p, t)x_e.
\end{align*}
The general solution of eq.~\eqref{eq:linearized} is
\begin{align}
    x(t) = \Phi(t)x_0 + \Phi(t)\int_0^t \Phi^{-1}(\tau)b(\tau)d\tau,\quad t\leq t_e,
    \label{eq:leqte}
\end{align}
where $\Phi(t) = \big(\varphi_1(t), \dots, \varphi_n(t)\big)$ is the matrix of linearly independent solutions $\varphi_i(t)$ of the homogeneous part of eq.~\eqref{eq:linearized}, i.e., $\forall i:\;\dot\varphi_i = A(p, t)\varphi_i$ with initial condition $\varphi_i(0) = {\rm e}_i$, the $i$'th unit vector.

First, we note that the original and linearized ODE's have the same sensitivity equations. The jumps of the sensitivities that we derive based on eq.~\eqref{eq:linearized} are therefore valid for the sensitivities of the original ODE, too. Second, the solution after the event time $t_e$ can be explicitly stated as
\begin{align}
    x(t) = \Phi(t-t_e)v_e + \Phi(t)\int_{t_e}^{t}\Phi^{-1}(\tau)b(\tau)d\tau,\quad t > t_e.
    \label{eq:gte}
\end{align}

\subsection{Sensitivities with respect to $t_e$}
Based on the explicit solutions, eqs.~\eqref{eq:leqte} and \eqref{eq:gte}, we find that
\begin{align}
    \left.\frac{\partial x}{\partial t_e}\right|_{t = t_e} =
    \left\{
    \begin{array}{ll}
    0     & \textrm{, for } t\nearrow t_e \\
    -A v_e + \frac{\partial v_e}{\partial t_e} - b(t_e)     & \textrm{, for }t \searrow t_e,
    \end{array}
    \right.
    \label{eq:jump}
\end{align}
where we have used that $\Phi(0) = \mathbb I$ is the unit matrix. The sensitivities with respect to $t_e$ jump at time point $t_e$. For $t > t_e$, the sensitivities propagate forward in time according to the sensitivity equations. The jump equation contains the derivative of $v_e$ which depends on the kind of the event.\\

\paragraph{Replacement}
The value of the $i$'th state, $x_i$, is set to a predefined value $v_{e, i}$. In that case, the derivative vanishes and the sensitivity at time $t_e$ is
\begin{align}
    \lim_{t\searrow t_e}\frac{\partial x_i}{\partial t_e} = \left.\frac{\partial f_i}{\partial x}\right|_{x_e}\cdot (x_e - v_e) - f_i(x_e),
    \label{eq:replace}
\end{align}
where we have used the definition of $b(t_e)$. In the equation, $\cdot: \mathbb R^n\times \mathbb R^n\rightarrow \mathbb R$ denotes the scalar product. While the state value $x_i$ jumps, the other states $x_j$, $j\neq i$, are continued. Continuation means that $v_{e, j}$ is set to $x_{e, j}$ and we need eq.~\eqref{eq:leqte} evaluated at $t_e$ to get $\frac{\partial v_e}{\partial t_e} = Ax_e + b(t_e)$. The sensitivities become
\begin{align}
    \lim_{t\searrow t_e}\frac{\partial x_j}{\partial t_e} = \left.\frac{\partial f_j}{\partial x}\right|_{x_e}\cdot (x_e - v_e).\label{eq:continued}
\end{align}
The contributions from $b$ cancel out. Due to the fact that all states except for $x_i$ are continued, the scalar product reduces to a single term. However, we find that an event at time $t_e$ affects sensitivities with respect to $t_e$ of both the affected and unaffected states.\\

\paragraph{Additive}
A constant $\Delta x_i$ is added to the value of the $i$'th state, $x_i$. In that case, $v_e$ is set to eq.~\eqref{eq:leqte} evaluated at $t_e$ plus the constant $\Delta x$. Therefore, the same argumentation as in the continued case holds for both, the affected and unaffected states, and the sensitivities are as eq.~\eqref{eq:continued} for states $i$ and $j\neq i$. \\

\paragraph{Multiplicative}
The value of the $i$'th state, $x_i$, is multiplied with a constant $\alpha_i$. Eq.~\eqref{eq:leqte} is evaluated at $t_e$ and multiplied by $\alpha$ to get $v_e$. Consequently, the derivative $\frac{\partial v_e}{\partial t_e}$ is computed and plugged into eq.~\eqref{eq:jump}, yielding
\begin{align}
    \lim_{t\searrow t_e}\frac{\partial x_i}{\partial t_e} = \left.\frac{\partial f_i}{\partial x}\right|_{x_e}\cdot (x_e - v_e) - (1-\alpha_i)f_i(x_e).
\end{align}
Same as for replacement events, the sensitivity of the affected and the unaffected states need to be distinguished. The term $(1-\alpha_i)f_i(x_e)$ only occurs for the affected state. Futhermore, all states except $x_i$ are continued. The scalar product therefore reduces to the $i$'th contribution.

@@

% Derive sensitivities with respect to the jump value
We use again the explicit solutions, eqs.~\eqref{eq:leqte} and \eqref{eq:gte}, and find that
\begin{align}
    \left.\frac{\partial x}{\partial v_e}\right|_{t = t_e} =
    \left\{
    \begin{array}{ll}
    0     & \textrm{, for } t\nearrow t_e \\
    \mathbb I     & \textrm{, for }t \searrow t_e.
    \end{array}
    \right.
    \label{eq:value}
\end{align}
Again, the sensitivities jump. They are forward propagated according to their sensitivity equations. Based on eq.~\eqref{eq:value}, we can construct sensitivities in case of additive events, where $v_e = x_e + \Delta x$ with $\Delta x \in \mathbb R^n$, and multiplicative events, where $v_{e, i} = \alpha_i x_{e, i}$ with $i = 1, \dots, n$. This is:
\begin{align}
    \left.\frac{\partial x}{\partial \Delta x}\right|_{t = t_e} = \mathbb I, \qquad \left.\frac{\partial x}{\partial \alpha}\right|_{t = t_e} = {\rm diag}(x_e).
\end{align}

>>>>

% Write abstract here
\begin{abstract}

  Forward sensitivity equations are frequently used in
  optimization. Based on forward sensitivities, the gradient and
  Hessian of the least squares function can be derived which allow to
  use gradient-based optimization methods. In the presence of events,
  i.e. sudden changes of the state variables, the differential
  equations and corresponding sensitivity equations are structurally
  unchanged. However, the events on states need to be accounted for as
  events in the sensitivities. Here, we derive these necessary events
  and present them in a way that helps with the implementation in
  computational software. Finally, the sensitivity events are
  illustrated on a typical example from the field of pharmacometrics.
\end{abstract}

@@

% Main Part
\section{Main Part}
\label{sec:main}

Let $\dot x = f(x, p, t)$ be a system of ordinary differential
equations (ODEs) with the states $x(t) \in \mathbb R^n$, parameters $p
\in \mathbb R^m$ and time $t \in \mathbb R$. The sensitivity equations
corresponding to this dynamic system are

\begin{align}
\begin{aligned}
    \frac{d}{dt}\frac{\partial x_i}{\partial p_j} &= \frac{d}{dp_j} f_i(x, p, t)\\
        & = \frac{\partial f_i}{\partial x_k}\frac{\partial x_k}{\partial p_j} +
            \frac{\partial f_i}{\partial p_j}.
\end{aligned}
 \label{eq:sens}
\end{align}

Let us assume that an event occurs at time $t_e$ that changes the
current state vector $x(t_e)$ to the value $v_e \in \mathbb R^n$. Both
$t_e$ and $v_e$ are assumed to be parameters for which sensitivities
are to be determined.

@@

The derivation is based on linearization of the ODE around the state
vector $x(t_e) = x_e$ at the event time.  This is the value of $x(t)$
at the time point just before the event occurs. With this choice,
the linearized ODE reads

\begin{align}
    \dot x \doteq A(p, t) x + b(p, t)
    \label{eq:linearized}
\end{align}

with

\begin{align*}
A(p, t) &= \left.\frac{\partial f(p, t)}{\partial x}\right|_{x_e}\\
b(p, t) &= f(x_e, p, t)- \left.\frac{\partial f(p, t)}{\partial x}\right|_{x_e}x_e.
\end{align*}

The general solution of eq.~\eqref{eq:linearized} is

\begin{align}
    x(t) = \Phi(t)x_0 + \Phi(t)\int_0^t \Phi^{-1}(\tau)b(\tau)d\tau,\quad t\leq t_e,
    \label{eq:leqte}
\end{align}

where $\Phi(t) = \big(\varphi_1(t), \dots, \varphi_n(t)\big)$ is the
matrix of linearly independent solutions $\varphi_i(t)$ of the
homogeneous part of eq.~\eqref{eq:linearized}, i.e., $\forall
i:\;\dot\varphi_i = A(p, t)\varphi_i$ with initial condition
$\varphi_i(0) = {\rm e}_i$, the $i$'th unit vector.

First, we note that the original and linearized ODE's have the same
sensitivity equations. The jumps of the sensitivities that we derive
based on eq.~\eqref{eq:linearized} are therefore valid for the
sensitivities of the original ODE, too. Second, the solution after the
event time $t_e$ can be explicitly stated as:

\begin{align}
    x(t) = \Phi(t-t_e)v_e + \Phi(t)\int_{t_e}^{t}\Phi^{-1}(\tau)b(\tau)d\tau,\quad t > t_e.
    \label{eq:gte}
\end{align}

\subsection{Sensitivities with respect to $t_e$}

Based on the explicit solutions, eqs.~\eqref{eq:leqte} and
\eqref{eq:gte}, we find that

\begin{align}
    \left.\frac{\partial x}{\partial t_e}\right|_{t = t_e} =
    \left\{
    \begin{array}{ll}
    0     & \textrm{, for } t\nearrow t_e \\ % Should we use t \geq t_e 
    -A v_e + \frac{\partial v_e}{\partial t_e} - b(t_e)     & \textrm{, for }t \searrow t_e,
    \end{array}
    \right.
    \label{eq:jump}
\end{align}

where we have used that $\Phi(0) = \mathbb I$ is the unit matrix. The
sensitivities with respect to $t_e$ jump at time point $t_e$. For $t >
t_e$, the sensitivities propagate forward in time according to the
sensitivity equations. The jump equation contains the derivative of
$v_e$ which depends on the kind of the event.\\

\paragraph{Replacement}

The value of the $i^{\textrm{th}}$ state, $x_i$, is set to a predefined value
$v_{e, i}$. In that case, the derivative vanishes and the sensitivity
at time $t_e$ is

\begin{align}
    \lim_{t\searrow t_e}\frac{\partial x_i}{\partial t_e} = \left.\frac{\partial f_i}{\partial x}\right|_{x_e}\cdot (x_e - v_e) - f_i(x_e),
    \label{eq:replace}
\end{align}

where we have used the definition of $b(t_e)$. In the equation,
$\cdot: \mathbb R^n\times \mathbb R^n\rightarrow \mathbb R$ denotes
the scalar product. While the state value $x_i$ jumps, the other
states $x_j$, $j\neq i$, are continued. Continuation means that $v_{e,
  j}$ is set to $x_{e, j}$ and we need eq.~\eqref{eq:leqte} evaluated
at $t_e$ to get $\frac{\partial v_e}{\partial t_e} = Ax_e +
b(t_e)$. The sensitivities become

\begin{align}
    \lim_{t\searrow t_e}\frac{\partial x_j}{\partial t_e} = \left.\frac{\partial f_j}{\partial x}\right|_{x_e}\cdot (x_e - v_e).\label{eq:continued}
\end{align}

The contributions from $b$ cancel out. Due to the fact that all states
except for $x_i$ are continued, the scalar product reduces to a single
term. However, we find that an event at time $t_e$ affects
sensitivities with respect to $t_e$ of both the affected and
unaffected states.\\

\paragraph{Additive}

A constant $\Delta x_i$ is added to the value of the $i$'th state,
$x_i$. In that case, $v_e$ is set to eq.~\eqref{eq:leqte} evaluated at
$t_e$ plus the constant $\Delta x$. Therefore, the same argumentation
as in the continued case holds for both, the affected and unaffected
states, and the sensitivities are as eq.~\eqref{eq:continued} for
states $i$ and $j\neq i$. \\

\paragraph{Multiplicative}

The value of the $i$'th state, $x_i$, is multiplied with a constant
$\alpha_i$. Eq.~\eqref{eq:leqte} is evaluated at $t_e$ and multiplied
by $\alpha$ to get $v_e$. Consequently, the derivative $\frac{\partial
  v_e}{\partial t_e}$ is computed and plugged into
eq.~\eqref{eq:jump}, yielding

\begin{align}
    \lim_{t\searrow t_e}\frac{\partial x_i}{\partial t_e} = \left.\frac{\partial f_i}{\partial x}\right|_{x_e}\cdot (x_e - v_e) - (1-\alpha_i)f_i(x_e).
\end{align}

Same as for replacement events, the sensitivity of the affected and
the unaffected states need to be distinguished. The term
$(1-\alpha_i)f_i(x_e)$ only occurs for the affected state. Furthermore,
all states except $x_i$ are continued. The scalar product therefore
reduces to the $i$'th contribution.

@@

% Derive sensitivities with respect to the jump value

We use again the explicit solutions, eqs.~\eqref{eq:leqte} and
\eqref{eq:gte}, and find that

\begin{align}
    \left.\frac{\partial x}{\partial v_e}\right|_{t = t_e} =
    \left\{
    \begin{array}{ll}
    0     & \textrm{, for } t\nearrow t_e \\
    \mathbb I     & \textrm{, for }t \searrow t_e.
    \end{array}
    \right.
    \label{eq:value}
\end{align}
Again, the sensitivities jump. They are forward propagated according to their sensitivity equations. Based on eq.~\eqref{eq:value}, we can construct sensitivities in case of additive events, where $v_e = x_e + \Delta x$ with $\Delta x \in \mathbb R^n$, and multiplicative events, where $v_{e, i} = \alpha_i x_{e, i}$ with $i = 1, \dots, n$. This is:
\begin{align}
    \left.\frac{\partial x}{\partial \Delta x}\right|_{t = t_e} = \mathbb I, \qquad \left.\frac{\partial x}{\partial \alpha}\right|_{t = t_e} = {\rm diag}(x_e).
\end{align}
